The Concurrent Calculi Formalisation Benchmark

### 2024-02-18 short-paper/short-paper.tex
more comments

\paragraph{Coinduction.} \todo{explain more what the challenge does.}
Coinduction is fundamental to establish properties of processes that
may never terminate, which is a typical situation.
Common uses for coinduction are to define strong and weak versions of
bisimulation, barbed notions of bisimulation, and prove trace equivalence. %,
%etc.
On the other hand, coinduction in proof assistants
is notoriously difficult: \lore{add here}
\cite{Castro-Perez2021} note that the most challenging
part of their mechanisation of multiparty session types
was the convoluted definition of infinite behaviours.

We base our challenge around bisimulation: a fundamental coinductively
established property for equating the
observed behaviour of potentially infinite processes. 
In this challenge, we study the notion of strong barbed bisimulation
(and bisimilarity, as in \cite{picalcbook})
over a fragment of the $\pi$-calculus and,
in particular, we focus on obtaining,
from strong barbed bisimilarity,
a congruence with respect to
parallel composition and substitution.

The introduction of bisimulation
is one of the first steps, when studying the
(observable) behaviour of process calculi, both
in classical textbooks (e.g., \cite{picalcbook})
and in almost all existent 
mechanisations, where we see a variety of
different formalisation choices.
As a significative example, we observe that,
while many authors rely
on the native support for coinduction offered by
proof assistants \cite{},
many others prefer a more set-theoretic
approach \cite{}.

@@

\subsection{Goals and how to contribute}
Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
%\alb{We oscillate between concurrency in general and process calculi, here and before}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

\subsubsection{Linearity.}
While a linear context is perhaps best seen as a multi-set, most proof assistants have better support for list processing.
The latter representation is intuitive, but may require establishing a large number of technical lemmata that are orthogonal to the mathematics of the problem under study, say proving type preservation for session types.
In this space, several designs are possible: one can label occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}) or \alb{something --- ask Francisco} as in \cite{Castro2020}.
Some authors instead impose a multiset structure over lists (\eg~\cite{ChaudhuriLR19,Danielsson12}); still, the effort required to develop the infrastructure is significant.

@@

For our challenge on linear reasoning we have chosen a type preservation theorem
for a system using session types.
% , whose use are common in concurrency theory.
This allows us to introduce linear reasoning with few definitions.
With our objective focused on linearity, we forego channel delegation
and opt for a reduction approach rather than employing a transition
system semantics.  Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names,n highlighting
their duality within the type system.  We have chosen a simple notion
of well-formedness, and do not  consider
deadlocks. While more sophisticated notions of well-formedness are
interesting, the proofs of these properties would distract from the
core linear aspects of the challenge.

>>>>

\paragraph{Coinduction.} \todo{explain more what the challenge does.}
Coinduction is fundamental to establish properties of processes that
may never terminate, which is a typical situation.
Common uses for coinduction are to define strong and weak versions of
bisimulation, barbed notions of bisimulation, prove trace equivalence,
etc.

We base our challenge around
bisimulation: a fundamental coinductively established property for equating the
observed behaviour of potentially infinite processes. 
In this challenge, we study the notion of strong barbed bisimulation
(and bisimilarity)
over a fragment of the $\pi$-calculus and,
in particular, we focus on obtaining,
from strong barbed bisimilarity,
a congruence with respect to
parallel composition and substitution.

The introduction of bisimulation
is one of the first steps, when studying the
semantics of process calculi, both
in classical textbooks (e.g., \cite{})
and in almost all existent 
mechanisations \cite{}
(where we see a variety
of different choices).

\cite{Castro-Perez2021} note that the most challenging part of their mechanisation was the convoluted definition of infinite behaviours.

@@

\subsection{Goals}% and how to contribute}
Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
%\alb{We oscillate between concurrency in general and process calculi, here and before}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

\subsubsection{Linearity.}

\alb{I would put the challenge before the literature}
For our challenge on linear reasoning we have chosen a type safety theorem
for a system using session types, combining as usual type preservation with absence of errors.
% , whose use are common in concurrency theory.
This allows us to introduce linear reasoning with few definitions.
With our objective focused on linearity, we forego channel delegation
and opt for a reduction approach rather than employing a transition
system semantics.  Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names, highlighting
their duality within the type system.  We have chosen a simple notion
of well-formedness, and do not  consider
deadlocks. While more sophisticated notions of well-formedness are
interesting, the proofs of these properties would distract from the
core linear aspects of the challenge.

While a linear context is perhaps best seen as a multi-set, most proof assistants have better support for list processing.
The latter representation is intuitive, but may require establishing a large number of technical lemmata that are orthogonal to the mathematics of the problem under study, say proving type preservation for session types.
In this space, several designs are possible: one can label occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}) or \alb{something --- ask Francisco} as in \cite{Castro2020}.
Some authors instead impose a multiset structure over lists (\eg~\cite{ChaudhuriLR19,Danielsson12}); still, the effort required to develop the infrastructure is significant.

@@
